---
title: "Marginal Causality"
author: "Gilles Monneret"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    fig_caption: yes
vignette: >
  %\VignetteIndexEntry{Marginal Causality}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r,echo=FALSE,message=FALSE}
require(knitr)
require(mvtnorm)
require(igraph)
require(Rgraphviz)
require(MarginalCausality)
ornon<-function(W,m=rep(0,ncol(W)),sigma2=rep(0.1,ncol(W))) # Obtention of the non oriented model parameters from the oriented model.
{
  d=length(m)
  if (abs(det(diag(d)-W))<10^(-8))
    {
      stop("Model not correctly specified")
    }
  L=solve(diag(d)-W)
  mu=matrix(m,1,d)%*%L
  Sigma=crossprod(L,diag(sigma2))%*%L
  return(list(mu=mu,Sigma=Sigma))
}
```
## Introduction and Method

  Distinguishing a real causal link between two variables is difficult, because of many possible confounders. The aim of this package is to choose whether or not a link between two variables is causal. To do so, two kinds of data are required: observational and interventional data for one particular variable of interest, for which we want to detect causal links with the other variables.
  
  We consider a gaussian structural equation model, as follows:
\begin{equation}
G=\mu_G+\varepsilon_G
\end{equation}
\begin{equation}
X=\alpha \times G+\mu_X+\varepsilon_X,
\end{equation}
where $\varepsilon$ are independent gaussian random variables. The observations are simulated under this model. 
```{r}
## Data clearly generated under a particular SEM: here, alpha=2, mu_G=-2 and mu_X=1
wt <- rnorm(n=20,mean=-2,sd=0.02)
wt <- cbind(wt,2*wt+rnorm(n=20,mean=1,sd=0.07))
```
For the interventions, the following model is used:
\begin{equation}
G=0
\end{equation}
\begin{equation}
X=\alpha \times G+\mu_X+\varepsilon_X.
\end{equation}
These interventions are critical: without them, in this setting, it is not possible to distinguish between a correlation and a causality model.
```{r}
## Interventional data
ko <- rnorm(n=20,mean=0,sd=10e-5)
ko <- cbind(ko,2*ko+rnorm(n=20,mean=1,sd=0.07))
```
The purpose of this package is to choose between a downstream causal model, where the node $G$ has a causal effect on $X$, and an alternative model, where $G$ does not have a causal effect on $X$: in this case, we can have a spurious relationship or a reverse causation. We have written the likelihood of these two models in the function *loglik.down* and *loglik.cor*. Using these two likelihood functions, we derive the posterior probability of the correlation model and the downstream model.
\[
\mathbb{P}(\text{cor}|\text{data})=
\frac{\mathbb{P}(\text{data}|\text{cor})\mathbb{P}(\text{cor})}
{\mathbb{P}(\text{data}|\text{cor})\mathbb{P}(\text{cor})
+\mathbb{P}(\text{data}|\text{down})\mathbb{P}(\text{down})}
\]
Then we can compute a Bayes factor, used to choose between the two models:
\[
B=\frac{\mathbb{P}(\text{cor}|\text{data})}{\mathbb{P}(\text{down}|\text{data})}
\]
The lower it is, the more we will likely choose the downstream model.

## Use of the package and simulation
Suppose that we have data simulated as: \newline
```{r,echo=FALSE,message=FALSE,fig.width=8,fig.height=6}
W=matrix(0,13,13) # On regarde 1,2,4,5,7,8,11,12,13
W[1:2,3]=c(0.1,-1.55)
W[2,4]=0.4
W[c(3,5),6]=c(1.5,-0.9)
W[2,7]=-1.1
W[5,8]=-2.2
W[6,9]=3
W[6:7,10]=c(-1.3,-0.2)
W[9,11]=-0.005
W[c(9,11),12]=c(1,0.1)
W[10,13]=1
colnames(W)=rownames(W)=1:13
L=solve(diag(13)-W)
k=1
s=k*exp(rnorm(13,0,0.5))
s[c(1,8,13)]=s[c(1,8,13)]*100
me=rnorm(13,0,2)
names(s)=names(me)=LETTERS[1:13]
par=list(mean=me,sigma2=s,W=W)
parobs=ornon(W = par$W,m = par$mean,sigma2 = par$sigma2)
Wtilde=W; Wtilde[,6]=0;
mtilde=par$mean; mtilde[6]=0;
sigma2tilde=par$sigma2; sigma2tilde[6]=0.01;
parko=ornon(W = Wtilde,m = mtilde,sigma2 = sigma2tilde)
set.seed(42)
N=48
g=graph.adjacency(W,weighted = TRUE)
gNEL=igraph.to.graphNEL(g)
eAttrs=c()
nAttrs=c()
nodcol=c(rep("blue",4),rep("green",3),rep("red",5),"yellow") ; 
names(nodcol)=nodes(gNEL)[c(1,2,3,5,4,7,8,9,10,11,12,13,6)] ;
nAttrs$fillcolor=nodcol
ew=as.character(unlist(edgeWeights(gNEL)))
ew=ew[setdiff(seq(along=ew),removedEdges(gNEL))]
names(ew)=edgeNames(gNEL)
eAttrs$label=ew
ewcol=ew; ewcol[1:14]="red"
eAttrs$fontcolor=ewcol
plot(gNEL,nodeAttrs=nAttrs,edgeAttrs=eAttrs,attrs=list(edge=list(fontsize=8)))
legend(x="bottomright",legend = c("Intervention","Upstream","Correlation","Downstream"),pch = 10, col=c("yellow","blue","green","red"))
```

> Figure 1: Simulation graph, with four states possibles.

An intervention was performed on node $6$ and we try to find on which nodes the node 6 has a downstream causal effect. For a specific set of parameters, we can calculate the log likelihood of the two models. Note that the function will return minus the log-likelihood.
```{r,fig.width=8,fig.height=6}
nrep=100
rdown=matrix(NA,nrep,13)
rcor=matrix(NA,nrep,13)
par=c(alpha=3,muG=11.42,muX=32,sigmaG=log(10),sigmaX=log(95))
for (rep in 1:nrep){
  # simulation
  xobs=rmvnorm(n = N/2,mean = parobs$mu,sigma = parobs$Sigma)
  xko=rmvnorm(n = N/2,mean = parko$mu,sigma = parko$Sigma)
  # loop on all tested genes
  for (j in c(1:13)[-6]){
    wt=xobs[,c(6,j)]
    ko=xko[,c(6,j)]
    rdown[rep,j]=loglik.down(wt=wt,ko=ko,par=par)
    rcor[rep,j]=loglik.cor(wt=wt,ko=ko,par=par)
  }
}

type=matrix(rep(1:13,nrep),nrep,13,byrow=TRUE)
boxplot(c(-rdown)~c(type),xlab="gene index",ylab="loglog-lik",col="red",ylim=c(min(-rcor,na.rm=TRUE)-5,max(-rdown,na.rm=TRUE)+5))
boxplot(c(-rcor)~c(type),xlab="gene index",ylab="loglog-lik",col="blue",add=TRUE)
```

> Figure 2: Log-likelihood, in blue the correlation model, in red the downstream model

To answer our question, one can use the Bayes factor defined before and based on these log-likelihoods.
```{r,fig.width=8,fig.height=6}
nrep=100
res=matrix(NA,nrep,13)
for (rep in 1:nrep){
  # simulation
  xobs=rmvnorm(n = N/2,mean = parobs$mu,sigma = parobs$Sigma)
  xko=rmvnorm(n = N/2,mean = parko$mu,sigma = parko$Sigma)
  # loop on all tested genes
  for (j in c(1:13)[-6]){
    wt=xobs[,c(6,j)]
    ko=xko[,c(6,j)]
    res[rep,j]=post.causal(wt = wt,ko= ko)$B
  }
}
type=matrix(rep(1:13,nrep),nrep,13,byrow=TRUE)
boxplot(c(log10(res))~c(type),xlab="gene index",ylab="log10(Bayes factor)",col=c(rep(1,7),rep(2,5)))
abline(v=7.5,lty=2,col="blue",lwd=2)
title("Boxplot of Bayes factor for all genes.")
```

> Figure 3: log10(BayesFactor) for each gene

Finally, we can compute the parameters for the downstream model for the red nodes, for the upstream model for the blue nodes and the correlation model for the green nodes. 

```{r}
N=48
nrep=25
coeff=vector(mode="list",length=nrep)
for (rep in 1:nrep){
  # simulation
  xobs=rmvnorm(n = N/2,mean = parobs$mu,sigma = parobs$Sigma)
  xko=rmvnorm(n = N/2,mean = parko$mu,sigma = parko$Sigma)
  # loop on all tested genes
  coeff[[rep]]=vector(mode="list",length=13)
  for (j in c(1:13)[-6]){
    wt=xobs[,c(6,j)]
    ko=xko[,c(6,j)]
    if (j>8) {
      coeff[[rep]][[j]]=post.causal(wt = wt,ko= ko)$down$par[1]
    } else if(match(j,c(4,7,8),nomatch = 0)>0) {
      coeff[[rep]][[j]]=undownstream(post.causal(wt = wt,ko= ko)$cor$par,to="cor")[1]
    } else {
      coeff[[rep]][[j]]=undownstream(post.causal(wt = wt,ko= ko)$cor$par,to="up")[1]
    }
  }
}
Effects <- matrix(unlist(coeff),nrow=nrep,ncol=12,byrow=TRUE)
trueCoeff=rep(NA,12)
trueCoeff[4]=parobs$Sigma[4,6]/sqrt(parobs$Sigma[4,4]*parobs$Sigma[6,6])
trueCoeff[6]=parobs$Sigma[7,6]/sqrt(parobs$Sigma[7,7]*parobs$Sigma[6,6])
trueCoeff[7]=parobs$Sigma[8,6]/sqrt(parobs$Sigma[8,8]*parobs$Sigma[6,6])
trueCoeff[c(1,2,3,5)]=L[c(1,2,3,5),6]
trueCoeff[c(8:12)]=L[6,9:13]
k_Effects <- rbind(trueCoeff,apply(Effects,2,mean),apply(Effects,2,sd))
rownames(k_Effects)=c("True Coefficient","Computed mean","Computed standard deviation")
kable(k_Effects,digits=3,col.names = paste("Node ",c(1:13)[-6],sep=""),caption="Several kinds of coefficients depending on the node: correlation coefficient for the nodes 4, 7 and 8, total upstream causal effect on 6 for nodes 1, 2, 3, 5 and total downstream causal effect of 6 for the other ones.")
```

> Table 1: Either a total effect or a correlation for each node relatively to node 6

\
\
\

This vignette was defined based on the paper **Identification of marginal causal relationships in
gene networks, from observational and
interventional expression data**, Gilles Monneret, Florence Jaffr\'ezic, Andrea Rau, Tatiana Zerjal, Gr\'egory Nuel, submitted to *BMC Bioinformatics*.
